---
title: "Deterministic Dose Escalation"
author: "Chase Mathis"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dose-Escalation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = TRUE,
  warning = FALSE
)
```

```{r load}
library(survivl)
options(digits = 4)
```

# Deterministic treatments

By default each treatment $A_k$ is *drawn* from a GLM given the past.
`dose_escalation()` instead makes $A_k$ a **deterministic function** of the past
-- a *dose-escalation* regime, where patients climb through ordered dose levels
but a bad side effect steps them back down. Drop it into the treatment slot of
`family` where a numeric code (e.g. `5`) would otherwise go.

# The escalation rule

For dose levels $0$ (placebo) to $L$ (`max_level`) and a contemporaneous
side-effect variable $Z_k$:

* **time 0:** coin flip, $A_0 \sim \mathrm{Bernoulli}(\texttt{init\_prob})$ on $\{0, 1\}$;
* **no side effect:** escalate, $A_k = \min(A_{k-1} + 1,\; L)$;
* **bad side effect:** de-escalate, $A_k = \max(A_{k-1} - 1,\; 0)$.

$Z_k$ is read from `paste0(side_effect, "_", k)`, simulated before $A_k$ each
step. The treatment formula's RHS is ignored (use `X ~ 1`) and needs no `beta`.

# A worked example

A study with a binary baseline `C`, a binary side-effect indicator `Z`, the
dose-escalation treatment `X` over levels $0,\dots,3$, and a Gamma outcome `Y`
on the cumulative dose.

```{r model}
T <- 7
L <- 3   # dose levels 0 (placebo), 1, 2, 3

formulas <- list(C ~ 1,
                 Z ~ X_l1 + C,                       # side effect; depends on previous dose
                 X ~ 1,                              # treatment: RHS ignored
                 Y ~ I(X_0 + X_1 + X_2 + X_3 + X_4 + X_5 + X_6) + C,
                 cop ~ 1)

family <- list(5,                                    # C binomial
               5,                                    # Z binomial (side-effect indicator)
               dose_escalation(max_level = L, side_effect = "Z", init_prob = 0.5),
               3,                                    # Y Gamma
               1)                                    # gaussian copula

pars <- list(C = list(beta = 0),
             Z = list(beta = c(-1/2, 0.4, 0.25)),    # intercept + X_{t-1} + C
             X = list(),                             # deterministic: no parameters
             Y = list(beta = c(0.05, 0.45, 0.05), phi = 1),
             cop = list(Y = list(Z = list(beta = 0.8472979))))

set.seed(1)
surv_model <- survivl_model(T = T, formulas = formulas, family = family, pars = pars)
dat <- rmsm(1e4, surv_model)
```

Doses shift upward over time but stay in $\{0,\dots,3\}$:

```{r dist}
A <- as.matrix(dat[, paste0("X_", 0:(T - 1))])
apply(A, 2, function(x) table(factor(x, levels = 0:L)))
```

Time 0 is a coin flip on placebo / lowest dose:

```{r coinflip}
table(A[, 1])
```

The rule holds exactly at every transition:

```{r check}
Z <- as.matrix(dat[, paste0("Z_", 0:(T - 1))])
all(sapply(2:T, function(k) {
  prev <- A[, k - 1]; bad <- Z[, k] > 0
  all(A[, k] == ifelse(bad, pmax(prev - 1, 0), pmin(prev + 1, L)))
}))
```

# Customising the side-effect trigger

For a non-binary side effect, pass a `trigger` mapping it to a logical, e.g.
de-escalate when a toxicity score exceeds a threshold:

```{r trigger, eval=FALSE}
dose_escalation(max_level = 3, side_effect = "Tox",
                trigger = function(z) z > 1.5)
```

# Recovering the causal effect with g-computation

The deterministic propensity $P(A_k \mid \text{past})$ is degenerate (0/1), so
IPW is unavailable; use the **g-formula**. Intervene on the dose path,
$\mathrm{do}(X_k = c)$, let `Z` evolve, and read off `Y`. A constant-dose
intervention is itself a deterministic rule returning the fixed level `c`:

```{r static-dose}
static_dose <- function(c) {
  structure(list(name = "deterministic",
                 rule = function(dat, k, vnm) rep(c, nrow(dat)),
                 pars = character(0), link = "identity", custom_links = NULL),
            class = c("causl_deterministic", "causl_family"))
}
```

Simulate at each dose level, pool, and fit the MSM:

```{r gcomp}
gcomp <- do.call(rbind, lapply(0:L, function(c) {
  fam_c <- family
  fam_c[[3]] <- static_dose(c)                       # do(X_k = c) for all k
  sm <- survivl_model(T = T, formulas = formulas, family = fam_c, pars = pars)
  d  <- rmsm(2e4, sm)
  data.frame(Y = d$Y, cumX = rowSums(d[, paste0("X_", 0:(T - 1))]), C = d$C)
}))

coef(glm(Y ~ cumX + C, data = gcomp, family = Gamma(link = "log")))
```

This recovers `pars$Y$beta` (`r paste(pars$Y$beta, collapse = ", ")`). A naive
fit to the observed data, without intervening, does not:

```{r naive}
dat$cumX <- rowSums(dat[, paste0("X_", 0:(T - 1))])
coef(glm(Y ~ cumX + C, data = dat, family = Gamma(link = "log")))
```

The dose coefficient is biased away from `r pars$Y$beta[2]`: `Z` is a
time-varying confounder (driven by the previous dose, coupled to `Y` via the
copula), which g-computation removes and the naive fit does not.

A future release will add a per-transition *slip* probability (the reserved
`slip` argument) to restore positivity for IPW.
