---
title: "Simpson's Paradox: When IPW and Naive Disagree on the Sign"
author: "Chase Mathis"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simpson-Flip}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
options(digits = 4)
```

# Goal

The `Longitudinal-Single-Outcome` vignette shows IPW and a naive outcome
regression giving roughly the same answer — it is hard to see which is
right just by looking at the numbers. This vignette goes the other way: we
build a longitudinal MSM whose **naive estimator gets the sign of the
treatment effect wrong**, while IPW correctly recovers the truth on average.
Same story as classical Simpson's paradox in cross-sectional data —
adjustment is not optional.

# The recipe

Three design choices make naive and IPW diverge dramatically.

1. **Drop the time-varying confounder $L$ out of the structural model.** All
   of $L \to Y$ flows through the copula, so the strength of the copula
   directly controls how much unmeasured confounding the naive model sees.
2. **Reverse $L \to A$ direction** (high $L$ → treated) and pair it with a
   **positive** $L$–$Y$ copula. Both bias channels now push naive in the
   *opposite* direction from the structural causal effect.
3. **Use a conditional copula**: the latent correlation
   $\rho_t = 2 \cdot \text{expit}(\beta_0 + \beta_1 A_{t-1}) - 1$ swings
   from $\approx 0.30$ when last period was untreated to $\approx 0.95$
   when it was. The naive marginal model cannot match this treatment-
   modified dependence.

We raise $T$ from 5 to 8 so that more periods are model-generated and
confounded (`dat0` fixes $A_0 = A_1 = A_2$, so only $A_3 \ldots A_{T-1}$ are
freshly drawn from the propensity model). The outcome $Y$ is Gaussian with
$\text{sd}(Y) = 20$.

```{r load}
library(survivl)

rho_to_beta <- function(rho) log(((rho + 1) / 2) / (1 - (rho + 1) / 2))
```

# DGP

```{r dgp-setup}
n     <- 5000
T_use <- 8

A_vars    <- paste0("A_", 0:(T_use - 1))
sumA_form <- as.formula(paste("Y ~ B + C + I(",
                              paste(A_vars, collapse = "+"), ")"))

make_init <- function(n) {
  qtls <- data.frame(matrix(runif(n * 4), ncol = 4))
  colnames(qtls) <- c("B", "C", "L_0", "A_0")
  qtls <- cbind(qtls, A_1 = qtls$A_0, A_2 = qtls$A_0)
  B   <- qbinom(qtls$B, 1, 0.1)
  C   <- qunif(qtls$C, 25, 35)
  L_0 <- qnorm(qtls$L_0, 0, 3)              # wide L, sd = 3
  A_0 <- A_1 <- A_2 <- qbinom(qtls$A_0, 1, 0.5)
  list(dat0 = cbind(B, C, L_0, A_0, A_1, A_2), qtls = qtls)
}

forms <- list(list(B ~ 1, C ~ 1),
              L ~ B + C,
              A ~ L_l0 + A_l1,                       # high L → treated
              sumA_form,
              list(Y = list(L ~ A_l1)))              # conditional copula
fams  <- list(list(5, 4), 1, 5, 1, 2)                # Y Gaussian, copula t

pars <- list(
  B = list(beta = 0),
  C = list(beta = 0),
  L = list(beta = c(0, 0, 0), phi = 9),
  A = list(beta = c(0, 0.8, 0.3)),                   # reversed L→A + persistence
  Y = list(beta = c(0, 0, 0, -0.8),                  # TRUTH: -0.8 per dose
           phi  = 400),                              # sd(Y) = 20
  cop = list(Y = list(L = list(
    beta = c(rho_to_beta(0.30), rho_to_beta(0.95) - rho_to_beta(0.30)),
    df   = 3,
    par2 = 3
  )))
)
```

# Naive and IPW in one helper

```{r estimators}
fit_one <- function(seed) {
  set.seed(seed)
  init <- make_init(n)
  sm   <- survivl_model(formulas = forms, family = fams, pars = pars,
                        T = T_use, dat = init$dat0, qtls = init$qtls)
  d    <- rmsm(n, sm)

  # IPW weights from the freshly-drawn periods t = 3, ..., T-1
  log_w <- rep(0, n)
  for (t in 3:(T_use - 1)) {
    a  <- paste0("A_", t); l <- paste0("L_", t); ap <- paste0("A_", t - 1)
    denom <- glm(as.formula(paste(a, "~", l, "+", ap)),
                 family = binomial(), data = d)
    numer <- glm(as.formula(paste(a, "~", ap)),
                 family = binomial(), data = d)
    pd <- predict(denom, type = "response")
    pn <- predict(numer, type = "response")
    aa <- d[[a]]
    log_w <- log_w + log(ifelse(aa == 1, pn, 1 - pn)) -
                     log(ifelse(aa == 1, pd, 1 - pd))
  }
  d$w <- exp(log_w)

  fit_naive <- lm(sumA_form, data = d)
  fit_ipw   <- lm(sumA_form, data = d, weights = w)
  trt       <- length(coef(fit_naive))
  c(naive = unname(coef(fit_naive)[trt]),
    ipw   = unname(coef(fit_ipw)[trt]))
}
```

# Monte Carlo

Per-replication IPW carries substantial variance under strong
$L \to A$ confounding because the propensity model becomes nearly
deterministic at the tails. We average over 20 replications:

```{r mc}
mc <- t(sapply(1:20, fit_one))
results <- data.frame(
  estimator    = c("Truth", "IPW", "Naive"),
  per_dose     = c(-0.8, mean(mc[, "ipw"]), mean(mc[, "naive"])),
  per_dose_sd  = c(NA,   sd(mc[, "ipw"]),   sd(mc[, "naive"])),
  cumulative_8 = c(-0.8, mean(mc[, "ipw"]), mean(mc[, "naive"])) * T_use
)
knitr::kable(results, digits = 3,
             caption = "Per-dose coefficient and total effect over 8 doses")
```

The story:

- **Truth:** each dose of treatment lowers $Y$ by 0.8 units; over 8 doses
  the cumulative effect is $-6.4$ ($\approx 0.3$ standard deviations of
  $Y$).
- **IPW** recovers a clearly negative effect, close to truth on average.
- **Naive** estimates a *positive* per-dose effect — it claims treatment
  *raises* $Y$. The signs of the two estimators are opposite.

# Why does this happen?

The copula gives high $L$ a strong positive association with $Y$ that is
not captured anywhere in the structural model. Because high $L$ also makes
treatment more likely (reversed $L \to A$), patients who get treated are
exactly those who would have had high $Y$ anyway. The naive regression
sees: people on treatment have high $Y$, so treatment must raise $Y$. IPW
corrects for the imbalance in $L$ between treated and untreated periods
and uncovers the true negative structural effect.

# Sanity check: null truth

If we set `pars$Y$beta[4] <- 0` and rerun, the naive estimator still
fabricates a positive harmful effect (about $+0.4$ per dose), while IPW
sits near zero. The naive bias is pure confounding, not a fact about
the treatment.
